---
title: "Mortality Trends"
author: "Steven J. Rigatti, MD, DBIM, FAAIM"
date: "2025-07-15"
output:
  ioslides_presentation:
    widescreen: true
    logo: "NewRRALogo.png"
    incremental: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(tidyverse)
library(readxl)
library(plotly)
library(RColorBrewer)

```



## Crude Rates in Farm Bureau States
#### High Income Counties

```{r crudeRates, fig.width=10, fig.algn='center'}
d <- read_excel("FBFS_COD_Research.xlsx", sheet = "SixGroupHighIncome")
d <- d %>% 
     select(Year, Sex, `Ten-Year Age Groups Code`, Deaths:MR_2019) %>% 
     filter(!is.na(Year)) %>% 
     mutate(Year = case_when(Year == "2024 (provisional)" ~ 2024,
                             Year == "2025 (provisional and partial)" ~ 2025,
                             TRUE ~ as.numeric(Year)),
            MR_2019 = round(MR_2019,2)) %>% 
     filter(Year <= 2024)
names(d) <- c("year","sex","age","deaths","population","crude_rate","q_2019", "MR")

## chart of crude rates by year, with lines colored by age group and sex in panels
p1 <- ggplot(d, aes(x = year, y = crude_rate, color = age)) +
     geom_line(lwd = 1) +
     facet_wrap(~sex) +
     labs(title = NULL,
          x = "Year",
          y = "Crude Rate") +
     theme_minimal() +
     scale_color_brewer(palette = "PuRd") +
     scale_y_log10() +
     # enlarge and darken the text
     theme(text = element_text(size = 16, color = "black"),
           legend.position = "bottom") +
     labs(color = "Age Group", x = NULL)
ggplotly(p1) %>%
      plotly::layout(legend=list(x=0, y=-0.2,
                                 xanchor='left',
                                 yanchor='bottom',
                                 orientation='h'))
```

## Rate Ratios vs. 2019 in Farm Bureau States
#### High Income Counties

```{r mr_2019, fig.width=10, fig.algn='center'}
dk_blues <- brewer.pal(name="Blues",n=9)[4:9]
p2 <- ggplot(d, aes(x = year, y = MR, color = age)) +
     geom_line(lwd = 1) +
     facet_wrap(~sex) +
     labs(title = "Mortality Rate Ratios vs. 2019 by Year and Age Group",
          x = "Year",
          y = "Rate Ratio (vs. 2019)") +
     theme_minimal() +
     scale_color_manual(values = dk_blues) +
     # scale_y_log10() +
     # enlarge and darken the text
     theme(text = element_text(size = 16, color = "black"),,
           legend.position = "bottom") +
     labs(color = "Age Group", 
          title = NULL, x = NULL)
ggplotly(p2) %>%
      plotly::layout(legend=list(x=0, y=-0.2,
                                 xanchor='left',
                                 yanchor='bottom',
                                 orientation='h'))
```


## Trends in FB States - EDR
```{r, fig.width=10, fig.algn='center'}
c <- read_excel("FBFS_COD_Research.xlsx", sheet = "provisionalData")
names(c) <- c("notes","year","yr","sex","sex_c","ucd_cause", "cause_code", "deaths", "population", "crude_rate", "rate", "q_2019", "MR", "EDR")

c <- c %>%
     filter(yr <= 2024,
            deaths >= 50) %>% 
     select(yr, sex, ucd_cause, cause_code, deaths, rate, q_2019, MR, EDR)

c <- c %>% 
     mutate(ucd_cause = ifelse(substr(ucd_cause, 1, 3) == "Acc", "Accidental Poisoning", ucd_cause))

#find which cause codes are in all of the years
causes <- c %>%
     group_by(cause_code) %>%
     filter(n() == 14) %>%
     ungroup() %>%
     select(cause_code) %>%
     distinct()

c <- c %>% filter(cause_code %in% causes$cause_code)
## change ucd_cause variable to include everything up to the first parenthesis
c$ucd_cause <- sub("\\(.*", "", c$ucd_cause)
c$MR <- round(c$MR, 2)
c$EDR <- round(c$EDR, 1)
p3 <- ggplot(c, aes(x = yr, y = EDR)) +
     geom_line(aes(color = sex)) +
     facet_wrap(~ucd_cause) +
     labs(title = NULL,
          x = NULL,
          y = NULL) +
     theme_minimal() +
     scale_color_brewer(palette = "Set1") +
     #scale_y_log10() +
     # enlarge and darken the text
     theme(legend.position = "none") +
     labs(color = "Cause of Death", x = NULL)
ggplotly(p3)

```

## Trends By Cause in FB States - MR
```{r,  fig.width=10, fig.algn='center'}
p4 <- ggplot(c, aes(x = yr, y = MR)) +
     geom_line(aes(color = sex)) +
     facet_wrap(~ucd_cause) +
     labs(title = "Mortality Ratio by Cause of Death",
          x = "Year",
          y = "Excess Death Rate") +
     theme_minimal() +
     scale_color_brewer(palette = "Set1") +
     #scale_y_log10() +
     # enlarge and darken the text
     theme(legend.position = "none") +
     labs(title = NULL,
          x = NULL,
          y = NULL)
ggplotly(p4)
```

## Alcoholic Liver Disease: Full US
#### Crude Rates Over Time
```{r, fig.width=10, fig.algn='center'}
alc <- read_xlsx("AlcoholLiverDisease.xlsx", sheet = "AlcoholLiverDisease")
alc <- alc[1:147,] #remove notes

alc <- alc %>% 
     filter(is.na(Notes))

names(alc) <- tolower(make.names(names(alc)))

alc <- alc %>% 
     select(ten.year.age.groups.code, sex, year.code:crude.rate)
names(alc) <- c("age_group", "sex", "year", "deaths", "population", "rate")

alc <- alc %>% 
     filter(year <= 2024) %>% 
     mutate(rate = round(deaths/population * 100000, 2))

alc1 <- alc %>% 
     filter(age_group %in% c("25-34", "35-44", "45-54", "55-64", "65-74", "75-84")) %>%
     ggplot(aes(x = year, y = rate, color = age_group)) +
     geom_line(lwd = 1) +
     facet_wrap(~sex) +
     theme_dark() +
     theme(text = element_text(size = 16, color = "black")) +
     scale_color_brewer(palette = "PuRd") +
     #scale_y_log10() +
     labs(x = NULL, y =NULL, color = "Age Group")

ggplotly(alc1) %>%
     plotly::layout(legend=list(x=0, y=-0.3,
                                xanchor='left',
                                yanchor='bottom',
                                orientation='h'))

```

## Alcoholic Liver Disease: Full US
#### Mortality Ratios vs. 2019
```{r, fig.width=10, fig.algn='center'}
q2019 <- alc %>% 
     filter(year == 2019) %>% 
     select(sex, age_group, rate) %>% 
     rename(rate_2019 = rate)

alc <- left_join(alc, q2019)
alc <- alc %>% 
     mutate(mr_2019 = round(rate /rate_2019, 2)) 

alc2 <- alc %>% 
     filter(age_group %in% c("25-34", "35-44", "45-54", "55-64", "65-74", "75-84")) %>%
     ggplot(aes(x = year, y = mr_2019, color = age_group)) +
     geom_line(lwd = 1) +
     facet_wrap(~sex) +
     theme_dark() +
     theme(text = element_text(size = 16, color = "black")) +
     scale_color_brewer(palette = "Blues") +
     #scale_y_log10() +
     labs(x = NULL, y =NULL, color = "Age Group")

ggplotly(alc2) %>%
     plotly::layout(legend=list(x=0, y=-0.3,
                                xanchor='left',
                                yanchor='bottom',
                                orientation='h'))
```

## Cirrhosis Mortality Trends: Full US
#### Crude Rates Over Time
```{r, fig.width=10, fig.algn='center'}
cir <- read_xlsx("cirrhosis.xlsx", sheet = "Cirrhosis")


cir <- cir[1:138,] #remove notes

cir <- cir %>% 
     filter(is.na(Notes))

names(cir) <- tolower(make.names(names(cir)))

cir <- cir %>% 
     select(ten.year.age.groups.code, sex, year.code:crude.rate)
names(cir) <- c("age_group", "sex", "year", "deaths", "population", "rate")

cir <- cir %>% 
     filter(year <= 2024) %>% 
     mutate(rate = round(deaths/population * 100000, 2))

cir1 <- cir %>% 
     filter(age_group %in% c("25-34", "35-44", "45-54", "55-64", "65-74", "75-84")) %>%
     ggplot(aes(x = year, y = rate, color = age_group)) +
     geom_line(lwd = 1) +
     facet_wrap(~sex) +
     theme_dark() +
     theme(text = element_text(size = 16, color = "black")) +
     scale_color_brewer(palette = "PuRd") +
     #scale_y_log10() +
     labs(x = NULL, y =NULL, color = "Age Group")

ggplotly(cir1) %>%
     plotly::layout(legend=list(x=0, y=-0.3,
                                xanchor='left',
                                yanchor='bottom',
                                orientation='h'))
```

## Cirrhosis Mortality Trends: Full US
#### Mortality Ratios vs. 2019
```{r, fig.width=10, fig.algn='center'}
q2019 <- cir %>% 
     filter(year == 2019) %>% 
     select(sex, age_group, rate) %>% 
     rename(rate_2019 = rate)

cir <- left_join(cir, q2019)
cir <- cir %>% 
     mutate(mr_2019 = round(rate /rate_2019, 2)) 

cir2 <- cir %>% 
     filter(age_group %in% c("25-34", "35-44", "45-54", "55-64", "65-74", "75-84")) %>%
     ggplot(aes(x = year, y = mr_2019, color = age_group)) +
     geom_line(lwd = 1) +
     facet_wrap(~sex) +
     theme_dark() +
     theme(text = element_text(size = 16, color = "black")) +
     scale_color_brewer(palette = "Blues") +
     #scale_y_log10() +
     labs(x = NULL, y =NULL, color = "Age Group")

ggplotly(cir2) %>%
     plotly::layout(legend=list(x=0, y=-0.3,
                                xanchor='left',
                                yanchor='bottom',
                                orientation='h'))

```

